An Algorithm For Spatial Mapping Using a Hexagon Tilegram, With Application to Australian Maps

### 2019-07-30 algorithmRjournal/algorithmRjournal.Rmd
Cleaned file

There are four tasks to create a hexagon tilegram.
These steps can be executed by the main function, `create_hexmap`, or can be applied separately for more flexibility.

@@

```{r libraries}
# To install sugarbag:
# devtools::install_github("srkobakian/sugarbag")
library(sugarbag)

library(tidyverse)
```

@@

The `create_hexmap` function requires several parameters, if they are not provided, the information will be derived from the polygon set used. If users already have a grid and set of centroids they may choose to only use the `allocate` function.

**The following must be provided to `create_hexmap`:**
```{r table_necessary, echo = FALSE}
knitr::kable(tibble::tibble(parameter = c('shp', 
  'sf_id',
  'focal_points'),
  description = c(
    'an sf object containing the polygon information', 
    'name of a column that distinguishes unique areas',
    'a data frame of reference locations used to allocate hexagons'
    )), format = "latex")
```

@@

The centroids can be derived from the set of polygons:

```{r derive_centroids}
## Use the create_centroids function to find polygon centroids:
centroids <- create_centroids(shp_sf = tas_lga, sf_id = "LGA_CODE16")
## Join to data for each area
centroids <- left_join(centroids,
  sf::st_drop_geometry(tas_lga), by = "LGA_CODE16")
```

```{r end_cents, fig.cap="Local Government Areas and derived centroids.", fig.height=2.5, fig.width = 5}
fort_lga <- fortify_sfc(tas_lga)
polys <- ggplot() + 
  geom_polygon(aes(x = long, y = lat, group = interaction(LGA_NAME16, polygon)), data = fort_lga, fill = "grey", size = 0.3, colour = "lightgrey") +theme_void() + coord_equal()

@@

```{r bbox}
## create the bounding box from the set of polygon centroids:
bbox <- tibble::tibble(min = c(min(centroids$longitude, na.rm = TRUE), min(centroids$latitude, na.rm = TRUE)),
        max = c(max(centroids$longitude, na.rm = TRUE), max(centroids$latitude, na.rm = TRUE)))

@@

A sequence is created for longitude columns, and another for latitude rows.

The sequences begin at the minimum longitude (latitude), minus the buffer distance.
Equally spaced intervals the size of the hexagons are created up to the maximum longitude (latitude), plus the buffer distance.

An individual point is created from all intersections of the longitude columns and latitude row sequences.

```{r grid}
grid <- tibble::as_tibble(
  expand.grid(
    hex_long = seq(bbox$min[1] - buffer_dist, # minimum
      bbox$max[1] + buffer_dist, # maximum
      hex_size), # distance between hexagons
    hex_lat = seq(bbox$min[2] - buffer_dist, # minimum
      bbox$max[2] + buffer_dist, # maximum
      hex_size))) # distance between hexagons
```

```{r plot_grid}
grid1 <- ggplot() + geom_polygon(aes(x = long, y = lat, group = interaction(LGA_NAME16, polygon)), data = fort_lga, fill = "grey", size = 0.3, colour = "lightgrey") + geom_point(aes(x=longitude, y = latitude), data= centroids, colour = "#b2df8a") + theme_void() + coord_equal() +
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#1f78b4", data = grid, size = 0.75) 
```

A square grid will not facilitate tessellated hexagons. Every second latitude row of points will be shifted right by half of the hexagon size.

```{r shift}
# Find every second latitude
shift_lat <- grid %>% dplyr::select(hex_lat) %>%
    dplyr::distinct() %>%
    dplyr::filter(dplyr::row_number() %% 2 == 1) %>% unlist()

# Shift the longitude of every second latitude to the right to make hex structure
grid <- grid %>%
    dplyr::mutate(hex_long = ifelse(hex_lat %in% shift_lat, hex_long,
        hex_long + (hex_size / 2))) %>%
    dplyr::mutate(id=1:NROW(.)) %>%
    dplyr::mutate(assigned=FALSE)
```

```{r plot_shift}
grid2 <- ggplot() + geom_polygon(aes(x = long, y = lat, group = interaction(LGA_NAME16, polygon)), data = fort_lga, fill = "grey", size = 0.3, colour = "lightgrey") + geom_point(aes(x=longitude, y = latitude), data = centroids, colour = "#b2df8a") + theme_void() + coord_equal() +
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#1f78b4", data = grid, size = 0.75) 
```

Each point is now given an ID.

```{r grid_id}
full_grid <- grid <- grid %>%
    mutate(hex_long_int = dense_rank(hex_long)-1,
        hex_lat_int = dense_rank(hex_lat)-1)
```

```{r, fig.cap="Grid points to create a tilegram.", fig.height=2.5, fig.width = 5}
gridExtra::grid.arrange(grid1, grid2, nrow = 1)
```

@@

```{r windows}
nlong <- length(unique(grid$hex_long))
nlat <- length(unique(grid$hex_lat))

@@

<!--
Figure of rolling window construction
-->

### Step 3: Filtering the grid

@@

```{r filter_grid, echo = FALSE}
    # APPLY A BUFFER
    # change buffer to amount of hexagons (ints) either side
    hex_buffer <- floor(buffer_dist/hex_size)

@@

```{r split_centroids}
# Split the centroid data set
centroids <- centroids %>% 
  nest(longitude, latitude) %>% 
  mutate(closest = map(data, closest_focal_point, focal_points = capital_cities %>% filter(points == "Hobart"))) %>% 
  unnest(data, closest) %>% 
  arrange(focal_distance)
```

@@

## Allocate each centroids to a hexagon grid point

@@

```{r centroids, eval = FALSE}
hexmap_allocation <- allocate(
  centroids = centroids,
  sf_id = "LGA_CODE16",
  hex_grid = grid,
  hex_size = 0.2, # same size used in create_grid
  hex_filter = 10,
  width = 35,
  focal_points = capital_cities,
  verbose = TRUE
)
```

@@

```{r arrange}
s_centroids <- centroids %>% arrange(focal_distance)

@@

```{r orig_dist}
hex_filter = 10
width = 35
# keep value to reset expanded distances
expand_dist <- hex_filter
```

@@

```{r available_points}
# filter for only the available hex grid points
hex_grid <- grid %>% filter(!assigned)
```

@@

```{r close_points}
filter_dist <- hex_filter*hex_size

@@

```{r buffer_circle}
    hex_grid <- hex_grid %>%
        rowwise %>%
        mutate(
        hex_lat_c = hex_lat - flat,
        hex_long_c = hex_long - flong) %>%
        mutate(hyp = ((hex_lat_c^2) + (hex_long_c^2))^(1/2))

@@

```{r buffers, fig.height=2.5, fig.width = 5}
gridExtra::grid.arrange(buff1, buff2, nrow=1)
```

@@

```{r buffer_slice}
# Filter for angle within circle

        if (hex_grid$angle_minus[1] < hex_grid$angle_plus[1]) {
            hex_grid <- hex_grid %>%
                # create slice of 60 degrees from centroid
                filter(angle_minus < hex_angle & hex_angle < angle_plus)
        } else {
            hex_grid <- hex_grid %>%
                # create slice of 60 degrees from centroid
                filter(hex_angle < angle_plus | angle_minus > hex_angle)
        }
```

```{r plot_slice}
ggplot() + geom_polygon(aes(x = long, y = lat, group = interaction(LGA_NAME16, polygon)), data = fort_lga, fill = "grey", size = 0.3, colour = "lightgrey") + theme_void() + coord_equal() +
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#a6cee3", data = grid, size = 0.25) +
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#1f78b4", data = hex_grid, size = 0.75) + 
  geom_point(aes(x=longitude, y = latitude), data = centroid1, colour = "#b2df8a", size = 0.8) 
```

@@

```{r choose_point}
# Choose first available point
cent <- centroid1 %>% dplyr::rename(focal_point = points, focal_dist = focal_distance, focal_angle = angle)

@@

```{r allocated}
hex_points_df <- centroid_allocation %>% 
  fortify_hexagon(hex_size = hex_size,
  sf_id = "LGA_CODE16")
```

```{r plot_allocated}
ggplot() + 
  geom_polygon(aes(x = long, y = lat, group = interaction(LGA_NAME16, polygon)), data = fort_lga %>% filter(LGA_CODE16=="62810"), fill = "grey", size = 0.3, colour = "lightgrey") +
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#1f78b4", data = hex_grid, size = 0.75) + 
  geom_point(aes(x=longitude1, y = latitude1), data = centroid1, colour = "#fb9a99", size = 1) + 
  geom_point(aes(x=longitude, y = latitude), data = centroid1, colour = "#b2df8a", size = 0.8) + 
  theme_void() +
  coord_equal()
```

@@

```{r full_hexmapcode, echo = TRUE}
# Create centroids set
centroids <- create_centroids(tas_lga, "LGA_CODE16")
# Create hexagon location grid
grid <- create_grid(centroids = centroids,
    hex_size = 0.3,
    buffer_dist = 1.2)
# Allocate polygon centroids to hexagon grid points
hex_allocated <- allocate(
  centroids = centroids,
  hex_grid = grid,
  sf_id = "LGA_CODE16",
  # same column used in create_centroids
  hex_size = 0.3,
  # same size used in create_grid
  hex_filter = 10,
  focal_points = capital_cities %>% filter(points == "Hobart"),
  width = 35,
  verbose = FALSE
)
```

```{r plot_final, fig.height=2.5, fig.width = 5}
h1 <- hex_allocated %>%
  fortify_hexagon(hex_size = 0.3, 
  sf_id = "LGA_CODE16") %>% 
  left_join(., tas_lga)

p1 <- fortify_sfc(tas_lga)
end_hex <- ggplot() +
  geom_polygon(data = p1, aes(x=long, lat, group = interaction(LGA_NAME16, polygon)), fill = "grey", size = 0.3, colour = "lightgrey") +
  geom_polygon(data = h1, aes(x=long, lat, group = LGA_CODE16)) + theme_void() + coord_equal()

>>>>

There are four tasks to create a hexagon tilegram.
These steps can be executed by the main function, `create_hexmap`, or can be implemented separately for more flexibility.

@@

```{r libraries, message=FALSE, warning=FALSE, comment = FALSE}
# To install sugarbag:
# devtools::install_github("srkobakian/sugarbag")
library(sugarbag)
library(tidyverse)
```

@@

The `create_hexmap` function requires several parameters, if they are not provided, the information will be derived from the \CRANpkg{sf} set of shapes used. If users already have a grid and set of centroids they may choose to only use the `allocate` function.

*The following must be provided to `create_hexmap`:*
```{r table_necessary, echo = FALSE}
knitr::kable(tibble::tibble(parameter = c('shp', 
  'sf_id',
  'focal_points'),
  description = c(
    'an sf object containing the polygon information', 
    'name of a column that distinguishes unique areas',
    'a data frame of reference locations used to allocate hexagons'
    )), format = "latex")
```

@@

The centroids can be derived from the set of polygons:
```{r cents, message=FALSE, warning=FALSE, comment = FALSE}
centroids <- create_centroids(shp_sf = tas_lga, sf_id = "LGA_CODE16")
```
```{r derive_centroids, echo = FALSE}
## Use the create_centroids function to find polygon centroids:
## Join to data for each area
centroids <- left_join(centroids,
  sf::st_drop_geometry(tas_lga), by = "LGA_CODE16")
```

```{r end_cents, fig.cap="Local Government Areas and derived centroids.", fig.height=2.5, fig.width = 5, echo = FALSE}
fort_lga <- fortify_sfc(tas_lga)
polys <- ggplot() + 
  geom_polygon(aes(x = long, y = lat, group = interaction(LGA_NAME16, polygon)), data = fort_lga, fill = "grey", size = 0.3, colour = "lightgrey") +theme_void() + coord_equal()

@@

```{r}
grid <- create_grid(centroids = centroids, hex_size = 0.2, buffer_dist = 1.2)
```

```{r bbox, echo = FALSE}
## create the bounding box from the set of polygon centroids:
bbox <- tibble::tibble(min = c(min(centroids$longitude, na.rm = TRUE), min(centroids$latitude, na.rm = TRUE)),
        max = c(max(centroids$longitude, na.rm = TRUE), max(centroids$latitude, na.rm = TRUE)))

@@

A sequence is created for longitude columns, and another for latitude rows.
The sequences begin at the minimum longitude (latitude), minus the buffer distance.
Equally spaced intervals the size of the hexagons are created up to the maximum longitude (latitude), plus the buffer distance.
An individual point is created from all intersections of the longitude columns and latitude row sequences.
```{r grid, echo = FALSE}
grid <- tibble::as_tibble(
  expand.grid(
    hex_long = seq(bbox$min[1] - buffer_dist, # minimum
      bbox$max[1] + buffer_dist, # maximum
      hex_size), # distance between hexagons
    hex_lat = seq(bbox$min[2] - buffer_dist, # minimum
      bbox$max[2] + buffer_dist, # maximum
      hex_size))) # distance between hexagons

grid1 <- ggplot() + geom_polygon(aes(x = long, y = lat, group = interaction(LGA_NAME16, polygon)), data = fort_lga, fill = "grey", size = 0.3, colour = "lightgrey") + geom_point(aes(x=longitude, y = latitude), data= centroids, colour = "#b2df8a") + theme_void() + coord_equal() +
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#1f78b4", data = grid, size = 0.75) 
```
A square grid will not facilitate tessellated hexagons. Every second latitude row of points will be shifted right by half of the hexagon size.
```{r shift, echo = FALSE}
# Find every second latitude
shift_lat <- grid %>% dplyr::select(hex_lat) %>%
    dplyr::distinct() %>%
    dplyr::filter(dplyr::row_number() %% 2 == 1) %>% unlist()

# Shift the longitude of every second latitude to the right to make hex structure
grid <- grid %>%
    dplyr::mutate(hex_long = ifelse(hex_lat %in% shift_lat, hex_long,
        hex_long + (hex_size / 2))) %>%
    dplyr::mutate(id=1:NROW(.)) %>%
    dplyr::mutate(assigned=FALSE)

grid2 <- ggplot() + geom_polygon(aes(x = long, y = lat, group = interaction(LGA_NAME16, polygon)), data = fort_lga, fill = "grey", size = 0.3, colour = "lightgrey") + geom_point(aes(x=longitude, y = latitude), data = centroids, colour = "#b2df8a") + theme_void() + coord_equal() +
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#1f78b4", data = grid, size = 0.75)

full_grid <- grid <- grid %>%
    mutate(hex_long_int = dense_rank(hex_long)-1,
        hex_lat_int = dense_rank(hex_lat)-1)
```

```{r, fig.cap="Grid points to create a tilegram.", fig.height=2.5, fig.width = 5, echo = FALSE}
gridExtra::grid.arrange(grid1, grid2, nrow = 1)
```

@@

```{r windows, echo = FALSE}
nlong <- length(unique(grid$hex_long))
nlat <- length(unique(grid$hex_lat))

@@

```
<!--
Figure of rolling window construction
-->
### Step 3: Filtering the grid

@@

```{r filter_grid, echo = FALSE, fig.height = 2.5, fig.width = 2.5}
    # APPLY A BUFFER
    # change buffer to amount of hexagons (ints) either side
    hex_buffer <- floor(buffer_dist/hex_size)

@@

```{r split_centroids, echo = FALSE}
# Split the centroid data set
centroids <- centroids %>% 
  nest(longitude, latitude) %>% 
  mutate(closest = map(data, closest_focal_point, focal_points = capital_cities %>% filter(points == "Hobart"))) %>% 
  unnest(data, closest) %>% 
  arrange(focal_distance)
```

@@

## Allocate each centroid to a hexagon grid point

@@

```{r centroids, eval = FALSE, fig.height = 2.5, fig.width = 2.5}
hexmap_allocation <- allocate(
  centroids = centroids,
  sf_id = "LGA_CODE16",
  hex_grid = grid,
  hex_size = 0.2, # same size used in create_grid
  hex_filter = 10,
  width = 35,
  focal_points = capital_cities,
  verbose = TRUE
)
```

@@

```{r arrange, echo = FALSE}
s_centroids <- centroids %>% arrange(focal_distance)

@@

```{r orig_dist, echo = FALSE}
hex_filter = 10
width = 35
# keep value to reset expanded distances
expand_dist <- hex_filter
```

@@

```{r available_points, echo = FALSE}
# filter for only the available hex grid points
hex_grid <- grid %>% filter(!assigned)
```

@@

```{r close_points, echo = FALSE}
filter_dist <- hex_filter*hex_size

@@

```{r buffer_circle, echo = FALSE}
    hex_grid <- hex_grid %>%
        rowwise %>%
        mutate(
        hex_lat_c = hex_lat - flat,
        hex_long_c = hex_long - flong) %>%
        mutate(hyp = ((hex_lat_c^2) + (hex_long_c^2))^(1/2))

@@

```{r buffers, echo = FALSE, fig.height=2.5, fig.width = 5}
gridExtra::grid.arrange(buff1, buff2, nrow=1)
```

@@

```{r buffer_slice, echo = FALSE}
# Filter for angle within circle

        if (hex_grid$angle_minus[1] < hex_grid$angle_plus[1]) {
            hex_grid <- hex_grid %>%
                # create slice of 60 degrees from centroid
                filter(angle_minus < hex_angle & hex_angle < angle_plus)
        } else {
            hex_grid <- hex_grid %>%
                # create slice of 60 degrees from centroid
                filter(hex_angle < angle_plus | angle_minus > hex_angle)
        }

buff3 <- ggplot() + geom_polygon(aes(x = long, y = lat, group = interaction(LGA_NAME16, polygon)), data = fort_lga, fill = "grey", size = 0.3, colour = "lightgrey") + theme_void() + coord_equal() +
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#a6cee3", data = grid, size = 0.25) +
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#1f78b4", data = hex_grid, size = 0.75) + 
  geom_point(aes(x=longitude, y = latitude), data = centroid1, colour = "#b2df8a", size = 0.8) 
```

@@

```{r choose_point, echo = FALSE}
# Choose first available point
cent <- centroid1 %>% dplyr::rename(focal_point = points, focal_dist = focal_distance, focal_angle = angle)

@@

```{r allocated, echo = FALSE}
hex_points_df <- centroid_allocation %>% 
  fortify_hexagon(hex_size = hex_size,
  sf_id = "LGA_CODE16")

buff4 <- ggplot() + 
  geom_polygon(aes(x = long, y = lat, group = interaction(LGA_NAME16, polygon)), data = fort_lga %>% filter(LGA_CODE16=="62810"), fill = "grey", size = 0.3, colour = "lightgrey") +
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#1f78b4", data = hex_grid, size = 0.75) + 
  geom_point(aes(x=longitude1, y = latitude1), data = centroid1, colour = "#fb9a99", size = 1) + 
  geom_point(aes(x=longitude, y = latitude), data = centroid1, colour = "#b2df8a", size = 0.8) + 
  theme_void() +
  coord_equal()
```

```{r buffs, fig.height = 2.5, fig.width = 5, echo = FALSE}
gridExtra::grid.arrange(buff3, buff4 , nrow=1)
```

@@

```{r full_hexmapcode, echo = TRUE, warning = FALSE}
# Create centroids set
centroids <- create_centroids(tas_lga, "LGA_CODE16")
# Create hexagon location grid
grid <- create_grid(centroids = centroids,
    hex_size = 0.3,
    buffer_dist = 1.2)
# Allocate polygon centroids to hexagon grid points
hex_allocated <- allocate(
  centroids = centroids,
  hex_grid = grid,
  sf_id = "LGA_CODE16",
  # same column used in create_centroids
  hex_size = 0.3,
  # same size used in create_grid
  hex_filter = 10,
  focal_points = capital_cities %>% filter(points == "Hobart"),
  width = 35,
  verbose = FALSE
)
```

# Transform the data for plotting

To plot polygons using \CRANpkg{ggplot} the \CRANpkg{sf} object can be transformed to a set of indidivual points that can be joined as a polygon.

```{r create_plot_final, fig.height=2.5, fig.width = 5, message = FALSE, warning = FALSE}
h1 <- hex_allocated %>%
  fortify_hexagon(hex_size = 0.3, 
  sf_id = "LGA_CODE16") %>% 
  left_join(., tas_lga)
```

```{r plot_final, echo = FALSE, fig.height=2.5, fig.width = 5, message = FALSE, warning = FALSE}
p1 <- fortify_sfc(tas_lga)
end_hex <- ggplot() +
  geom_polygon(data = p1, aes(x=long, lat, group = interaction(LGA_NAME16, polygon)), fill = "grey", size = 0.3, colour = "lightgrey") +
  geom_polygon(data = h1, aes(x=long, lat, group = LGA_CODE16)) + theme_void() + coord_equal()
